%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%%  advanced.tex         IRREDSOL documentation           Burkhard H\"ofling
%%
%%  @(#)$Id$
%%
%%  Copyright (C) 2003 by Burkhard H\"ofling, 
%%  Institut f\"ur Geometrie, Algebra und Diskrete Mathematik
%%  Technische Universit\"at Braunschweig, Germany
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Chapter{Advanced information}

This chapter gives some additional information about the {\IRREDSOL} library.
It describes details about the library design (see "Design of the group library")
and some advanced functions for managing data in the GAP workspace, which may be
useful if you run out of memory. See "Loading and unloading group data manually"
and "Loading and unloading recognition data manually".

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Design of the group library}

To avoid redundancy, the package {\IRREDSOL} does not store 
lists of irreducible subgroups of $GL(<n>, <q>)$ but
only has lists ${\cal
A}_{n,q}$ of subgroups of $GL(<n>, <q>)$ such that

\beginlist
\item{--} each group in ${\cal A}_{n,q}$ is absolutely irreducible and
solvable

\item{--} ${\cal A}_{n,q}$ contains a conjugate of each absolutely
irreducible solvable subgroup of

\item{--} no two groups in ${\cal A}_{n,q}$ are conjugate

\item{--} no conjugate of $G$ is a subgroup of $GL(n, q_0)$, where
$GF(q_0)$ is a proper subfield of $GF(q)$.

\endlist Here, ``conjugate'' means ``conjugate in $GL(n,q)$''. We will
briefly say that ${\cal A}_{n,q}$ contains, up to conjugacy, all absolutely
irreducible solvable subgroups of $GL(n,q)$ which cannot be written over a
proper subfield.

These lists are sufficient to reconstruct lists of irreducible soluble
subgroups of $GL(n, q)$, since  any such subgroup can be obtained from an
absolutely irreducible subgroup of $GL(n/d, q^d)$, where $d$ divides $n$,
by regarding the underlying $\F_{q^d}$-vector space as an $\F_q$-vector
space. Note that two subgroups of $GL(n, q)$ constructed in that way are
conjugate if, and only if, the images of the corresponding subgroups of
under Galois automorphisms of $\F_{q^d}/\F_q$ are conjugate in $GL(n/d, q^d)$.
This information also forms part of the {\IRREDSOL} library.

Note that by the Deuring-Noether theorem, two subgroups of $GL(n, q)$ are
conjugate in $GL(n, q)$ if, and only if, they are conjugate in $GL(n, q^d)$
for some $d > 1$. This ensures that lists of irreducible subgroups obtained
from the ${\cal A}_{n,q}$ do not contain conjugate subgroups. 


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Loading and unloading group data manually}

The functions in this section are not required for normal use of the data library.
They are only relevant if timing or conservation of memory is an issue.
\index{workspace!running out of}

\>LoadAbsolutelyIrreducibleSolvableGroupData(<n>, <q>) F

This function loads the data for $GL(<n>, <q>)$ into memory and does some pre-processing. If the data
is already loaded, the function does nothing. This function is called automatically when you access the
{\IRREDSOL} library, so most users  will not need this function.

\>`LoadedAbsolutelyIrreducibleSolvableGroupData()'%
{LoadedAbsolutelyIrreducibleSolvableGroupData}%
@{`LoadedAbsolutelyIrreducibleSolvable\\GroupData'} F

This function returns a list. Each entry consists of an integer <n> and a set <l>. The set
<l> contains all prime powers <q> such that the group data for $GL(<n>, <q>)$ is currently in memory.

\>`UnloadAbsolutelydIrreducibleSolvableGroupData([n [,q]])'%
{UnloadAbsolutelydIrreducibleSolvableGroupData}%
@{`UnloadAbsolutelydIrreducibleSolvable\\GroupData'} F

This function can be used to delete data for irreducible groups from the GAP workspace. If no argument
is given, all data will be deleted. If only <n> is given, all data for degree <n> (and any <q>) will
be deleted. If <n> and <q> are given, only the data for $GL(n, q)$ will be deleted from the GAP
workspace. Use this function if you run out of GAP workspace. The
data is automatically re-loaded when required.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Loading and unloading recognition data manually}

The functions in this section are not required for normal use of the data library.
They are only relevant if timing or conservation of memory is an issue.
\index{workspace!running out of}

\>`LoadedAbsolutelyIrreducibleSolvableGroupFingerprints()'%
{LoadedAbsolutelyIrreducibleSolvableGroupFingerprints}%
@{`LoadedAbsolutelyIrreducibleSolvableGroup\\Fingerprints'} F

This function returns a list. Each entry consists of an integer <n> and a set <l>. The set
<l> contains all prime powers <q> such that the recognition data for $GL(<n>, <q>)$ is currently in
memory.

\>`UnloadAbsolutelyIrreducibleSolvableGroupFingerprints([n [,q]])'%
{UnloadAbsolutelyIrreducibleSolvableGroupFingerprints}%
@{`UnloadAbsolutelyIrreducibleSolvableGroup\\Fingerprints'} F

This function can be used to delete recognition for irreducible groups from the GAP workspace. If no
argument is given, all data will be deleted. If only <n> is given, all data for degree <n> (and any
<q>) will be deleted. If <n> and <q> are given, only the data for $GL(n, q)$ will be deleted from the
GAP workspace. Use this function if you run out of GAP workspace. The
data is automatically re-loaded when required.





%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%E  
%%
